---
title: "Stat 240 Project Draft"
author: "Mai Tah Lee, Jon Chang, Sophie Schrag, Michitake Kuwabara"
date: "2024-12-05"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	error = TRUE,
	message = FALSE,
	warning = FALSE,
	include = FALSE
)
library(tidyverse)
theme_set(theme_minimal())
```

```{r include=FALSE}
data = read_csv("drug_consumption.csv")
```

# Stat 240 Final Project

## Introduction
Drug usage and overdose are critical public health challenges, nearly less than 100,000 people die annually from drug overdoses in the United States(cite). The patterns of drug consumption across different age groups can help us better understand the context behind these numbers, which could further guide prevention and intervention efforts. This analysis explores how age groups correlates with the usage of various drugs, both legal (e.g., alcohol) and illegal (e.g., cannabis, ecstasy) and what the trend in different drug consumption proportions across different age groups. By exploring these relationships, we aim to identify which drugs are most commonly used within each age group and examine the characteristics associated with these trends. Ultimately, this analysis finds that the most consumed drugs were caffeine, chocolate (which can cause abnormal behaviors and other sensations that can be considered an addictive substance[cite]), and alcohol among individuals aged 18-24, and overall drug consumption decreases significantly with increased age.


## Background

In order to answer our question, we need to know how much alcohol is being consumed by each age group from age 18 to 65 and up. The age groups consist of 18-24, 24-35, 35-44, 45-54, 55-64 and 65+. Our data set consists of data from one source that had this information 

#### Age and Drug Use Data(Drug_consumption.csv)
Our data came from a Drug consumption dataset(site source). This dataset had information on many different demographics and their consumption of various drug types, some legal and some illegal. The raw data contains many columns and 1885 rows (individuals); however for this study we only used columns that corresponded to age group (Age) and alcohol consumption (Alcohol). The Age column quantified the ages of all individuals so that each row or individual had a value across a broad range between -0.95 and 2.59,  that would put them in a certain age group. For example the value -0.95197 puts that individual into the age group 18-24 where 1.09449 puts them into 45-59. The Alcohol column placed each individual into one of 7 classes based on the frequency of their consumption. The classes included, CLO: never used; CL1: used over a decade ago, CL2: used in the last decade; CL3: used in the last year; CL4: used in the last month; CL5: used in the last week; and CL6 used in the last day. 

#### Combined Data
These data can be joined together in a separate table, grouping individuals by their respective age group, and then a statistical analysis can be conducted to see if alcohol consumption varies among age groups. 

```{r}
# Add a graph Here
```

## Analysis
#### Initial Impressions


