---
title: "projectDraft"
author: "Mai Tah Lee"
date: "2024-12-02"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
library(tidyverse)
theme_set(theme_minimal())
```

```{r}
data = read_csv("drug_consumption.csv")
head(data, n = 3)
```

# Prior Notes: 
  * Variables are standardized numerically by variance of values in the dataset. 
  * For readability They need to be reinitialized with their actual values. 

# Preprocessing

* 20241204-MTL:
* I'm going to first look at the distributions of what drug each person consumes.
* I'm also going to make a graph so that we can visually see the distribution of types of drug consumption between different persons. 
* Then I'll attempt to run other analysis as necessary. 

## Initialize Dataset for Readability and Other Uses

```{r}
# Saved list of age ranges for later use
age = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")
cat("age_range_variance=",length(age), "\n")

# Save list of drugs for later use
drugs = c("Alcohol", "Amphet", "Amyl", "Benzos", "Caff", "Cannabis", "Choc", "Coke", "Crack", "Ecstasy", "Heroin", "Ketamine", "Legalh", "LSD", "Meth","Mushrooms", "Nicotine", "Semer", "VSA")

cat("total_drugs=",length(drugs))
```

```{r}
# 20241204-MTL: I'm also going to clean the data into a new variable separate from
#               the original data (so that the original data will still be
#               accessible), it may be unnecessary and the process may be time
#               consuming, but it will make it more readable for other uses if
#               needed. It's better to have clean readable data vs numerical values
#               that make it harder to distinguish its value in the data.  


# ********************** Initialize Values **************************
Data = data %>% 
  mutate(
    
### Initialize Age
    Age = case_when(
      Age == -0.95197 ~ "18-24",
      Age == -0.07854 ~ "25-34",
      Age == 0.49788 ~ "35-44",
      Age == 1.09449 ~ "45-54", 
      Age == 1.82213 ~ "55-64", 
      Age == 2.59171 ~ "65+"
    ),
    
### Make Gender Binary (clean the data); 
# **** 0==Male(-0.48246), 1==Female (0.48246) 
    Gender = case_when(
      Gender >= 0 ~ 1, # female
      Gender < 0 ~ 0   # male
    ),
    
### Initialize Education
    Education = case_when(
      Education == -2.43591 ~ "Left School Before 16 years", 
      Education == -1.73790 ~ "Left School at 16 years", 
      Education == -1.43719 ~ "Left School at 17 years", 
      Education == -1.22751 ~ "Left School at 18 years", 
      Education == -0.61113 ~ "Some College,No Certificate Or Degree", 
      Education == -0.05921 ~ "Professional Certificate/ Diploma", 
      Education == 0.45468 ~ "University Degree", 
      Education == 1.16365 ~ "Masters Degree", 
      Education == 1.98437 ~ "Doctorate Degree", 
    ),

### Initialize Country
    Country = case_when(
      Country == -0.09765 ~"Australia",
      Country == 0.24923 ~"Canada",
      Country == -0.46841 ~"New Zealand",
      Country == -0.28519 ~"Other",
      Country == 0.21128 ~"Republic of Ireland",
      Country == 0.96082 ~"UK",
      Country == -0.57009 ~"USA"
    ),

### Initialize Ethnicity
    Ethnicity = case_when(
      Ethnicity == -0.50212 ~ "Asian",
      Ethnicity == -1.10702 ~ "Black",
      Ethnicity == 1.90725 ~ "Mixed-Black/Asian",
      Ethnicity == 0.12600 ~ "Mixed-White/Asian",
      Ethnicity == -0.22166 ~ "Mixed-White/Black",
      Ethnicity == 0.11440 ~ "Other",
      Ethnicity == -0.31685 ~ "White"
    )

# The Behavior score values will not be initialized, their numbers
# are standardized based on overall variance. (It will also be time consuming to relabel the values.)

  )

head(Data, 5)

```
```{r}
### Initialize Consumption Values
# *** Consumption is based on a rating scale from 0-6
# *** @JonChang; 20241204-MTL:  I see that you've added a drug usage in your code
#                               which regards those who selected > (greater than) 3 as 1: regular usage
#                               and those < (less than) 3 are encoded as 0: non-regular. The non-regular
#                               is a bit vague, as non-regular could mean more than the norm or less than
#                               the norm. In this case, it would be better to just say less-regular, I'm
#                               assuming that you plan to create a usage column to asses each individual
#                               persons' consumption of a drug.

# Consumption values will be encoded as 0:less-regular, 1:regular. 
# Note a categorical values isn't truly necessary, also if the values are numerical
# it simplifies coding the data. (you can just think of 0 as not regular, and 1 as regular)
Data <- Data %>% 
  mutate(
    across(all_of(drugs), ~ case_when(     
      . == "CL0" ~ 0,
      . == "CL1" ~ 0,
      . == "CL2" ~ 0,
      . == "CL3" ~ 0,
      . == "CL4" ~ 1,
      . == "CL5" ~ 1,
      . == "CL6" ~ 1
    ))
  ) %>% select( # removing the behavioral scores
    -Nscore,
    -Escore, 
    -Oscore,
    -Ascore,
    -Cscore,
    -Impulsive,
    -SS
  )
head(Data, 10)

```

# Proportion of Drug Consumption

## Proportions of Drug Use Across All Age Groups

```{r}

nTotal = nrow(Data) # total amount of rows or observations of persons
cat("observations=", nTotal, "\n")  # should be 1885 obs.

usage_prop = c() # list the the proportions of each drug used within the dataset
for (val in drugs) {
  drugSum = sum(Data[[val]])
  drugProp = drugSum / nTotal
  usage_prop = append(usage_prop, drugProp)
}

cat("usage_prop_variance=", length(usage_prop), "\n\n") # should expect to have 19 proportions
cat("usage_prop_contents=\n", usage_prop)
```

```{r}
# The overall proportions across all age groups
drugProps = tibble(
  Drugs = drugs,
  usageProp = usage_prop
) %>% arrange(desc(usageProp))
drugProps
```



```{r}
# The bar plot of the proportions of drug usage across all age groups
ggplot(drugProps, aes(x = Drugs, y = usageProp)) + 
  geom_bar(stat="identity",fill='blue', alpha=0.5)+
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(
    title="Proportion of Drug Consumption Across All Age Groups",
    x = "Drug Consumed",
    y = "Proportion of Drug Consumption"
  )
  
```
> Accross our data the most used drugs are (1) chocolate, (2), caffiene, and (3) alcohol. 


## Proportions of Drug Use by Different Age Groups



```{r}
#It's best to filter the data by age group then take the proportions
ageGroups_Use = list() 
nTotals = c()          

for (val in age) {
  ageDrug <- Data %>% 
    filter(Age == val)
  
  ageGroups_Use[[as.character(val)]] = ageDrug
  
  n = nrow(ageDrug)
  nTotals = append(nTotals, n)
}

# Check variances
cat("nTotals_Variance=", length(nTotals), "\n")       # Number of unique age groups
cat("ageGroups_Use_Variance=", length(ageGroups_Use), "\n") # Total rows

# create list containing bar plots
barPlots = list()

```

* In total there are 6 different age groups, which requires assessing the each of the drug usage proportions among their group.
* In total there should be 6 barplots to create, as well as 6 smooth plots to compare. We can also test different plots for better visuals. 


```{r}
# create a loop that can:
# 1. create a dataframe with all the different proportions of drug usage on drug type
# 2. create plots from the dataframes
ageGroup_Prop_DFS = list() # will contain a data frame consisting of the drug consumption proportions by age group

barPlot = list()

for (i in seq(1,6)) {
  group = ageGroups_Use[[i]] # should be the data frame of the current age group, excludes other age groups
  currAge = age[i] # should be the string value; for example "18-24"
  n = nrow(group) # the entire observations within each age group
  usageProp = NULL
  
  
  usage_prop = c()
  for (val in drugs) {
    drugSum = sum(group[[val]])
    drugProp = drugSum / n
    usage_prop = append(usage_prop, drugProp)
  }
  
  prop_df = tibble(
    Drugs = drugs,
    usageProp = usage_prop 
  ) 
  ageGroup_Prop_DFS[[length(ageGroup_Prop_DFS) + 1]] = prop_df
  
  
  plot = ggplot(prop_df, aes(x = Drugs, y = usageProp)) +
    geom_bar(stat="identity",fill='blue', alpha=0.5)+
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(
    title=sprintf("Proportion of Drug Consumption Across the Age Group %s",  currAge),
    x = "Drug Consumed",
    y = "Proportion of Drug Consumption"
  )
  barPlot[[length(barPlot) + 1]] = plot
}


```

## The Data Frame of Proportions of Drug Consumption of each Age Group

```{r}
# The different data frames of each usageProp

drugProps_df = tibble(
    Drugs = drugs,
    usageProp1 = ageGroup_Prop_DFS[[1]]$usageProp,
    usageProp2 = ageGroup_Prop_DFS[[2]]$usageProp,
    usageProp3 = ageGroup_Prop_DFS[[3]]$usageProp,
    usageProp4 = ageGroup_Prop_DFS[[4]]$usageProp,
    usageProp5 = ageGroup_Prop_DFS[[5]]$usageProp,
    usageProp6 = ageGroup_Prop_DFS[[6]]$usageProp
  ) 

drugProps_df

```

## Arranged Drug Consumption Proportion by Group

```{r}
for (i in seq(1, 6)) {
  df = ageGroup_Prop_DFS[[i]] %>% arrange(desc(usageProp))
  cat("Group=", i, "\n")
  print(df)
  cat("\n\n")
}

```

## Bar Plots of Drug Consumption Proprotion by Age Group

```{r}
for (i in seq(1, 6)) {
  plot = barPlot[[i]]
  print(plot)
}
```


> For the most part, from looking at all these bar graphs, we can see that the most consistent drug usage would be caffiene, chocolate and Alcohol. I would say caffiene and chocolate aren't as harmful compared
to other drugs like alcohol. Therefore, are focus could be simply just alcohol. 

# Proportion of Drug Consumption by Age Group

## Alcohol Usage Across Age groups

* Setting up alcohol proportions by age group

```{r}

alc = drugProps_df %>% filter(Drugs == "Alcohol")
alc = t(alc)
alc = alc[-1]

alc_prop = tibble(
  Age_Group = age
  ) %>% mutate(
    Alcohol_Prop = as.numeric(alc)
  )

alc_prop
```

```{r}
# plot the alcohol consumption proportion by age group
ggplot(alc_prop, aes(x = Age_Group, y = Alcohol_Prop, group = 1)) + 
  geom_line(aes(color = "Alcohol Proportion"), size = 1) + 
  geom_point(aes(color = "Alcohol Proportion"), size = 3) + 
  geom_smooth(
    method = "lm",
    aes(color = "Trend Line"), 
    size = 1
  ) +
  scale_color_manual(
    values = c("Alcohol Proportion" = "blue", "Trend Line" = "red"),
    name = "Legend"
  ) +
  labs(
    title = "Alcohol Consumption Proportion by Age Group",
    x = "Age Group",
    y = "Proportion of Alcohol Consumption"
  ) + 
  theme_minimal()
```

> This is an obious downward trend in drug consumption as age group increases. 

## Amphet Usage Across Age groups

```{r}

amp = drugProps_df %>% filter(Drugs == "Amphet")
amp = t(amp)
amp = amp[-1]

amp_prop = tibble(
  Age_Group = age
  ) %>% mutate(
    Amphet_Prop = as.numeric(amp)
  )

amp_prop
```

```{r}
# plot amphet usage by age group
ggplot(amp_prop, aes(x = Age_Group, y = Amphet_Prop, group = 1)) + 
  geom_line(aes(color = "Amphet Proportion"), size = 1) + 
  geom_point(aes(color = "Amphet Proportion"), size = 3) + 
  geom_smooth(
    method = "lm",
    aes(color = "Trend Line"), 
    size = 1
  ) +
  scale_color_manual(
    values = c("Amphet Proportion" = "blue", "Trend Line" = "red"),
    name = "Legend"
  ) +
  labs(
    title = "Amphet Consumption Proportion by Age Group",
    x = "Age Group",
    y = "Proportion of Amphet Consumption"
  ) + 
  theme_minimal()

```

* create loop to automate plotting for all drugs

```{r}
drug_plots = list()

for (drug in drugs) {
  # Extract the proportion of the current drug across age groups
  drug_data = drugProps_df %>% filter(Drugs == drug)
  drug_data = t(drug_data)  # Transpose the data
  drug_data = drug_data[-1]  # Remove the header row
  
  # Create data frame for the current drug
  drug_prop = tibble(
    Age_Group = age
  ) %>% mutate(
    Proportion = as.numeric(drug_data)
  )
  
  # Create the plot
  plot = ggplot(drug_prop, aes(x = Age_Group, y = Proportion, group = 1)) +
    geom_line(aes(color = "Drug Proportion"), size = 1) +  # Use static label for legend
    geom_point(aes(color = "Drug Proportion"), size = 3) +
    geom_smooth(
      method = "lm",
      aes(color = "Trend Line"), 
      size = 1
    ) +
    scale_color_manual(
      values = c(
        "Drug Proportion" = "blue", 
        "Trend Line" = "red"
      ),
      name = "Legend"
    ) +
    labs(
      title = paste(drug, "Consumption Proportion by Age Group"),
      x = "Age Group",
      y = paste("Proportion of", drug, "Consumption")
    ) + 
    theme_minimal()
  
  # Add the plot to the list
  drug_plots[[drug]] = plot
}
```

## All Plots for Proportions of Specific Drug Consumption by Age group

```{r}
for (drug in drugs) {
  print(drug_plots[[drug]])
}

```

> From all these graphs, the correlation with age group is that as age increases, drug consumption decreases, this is very intuitive and and obvious trend. 


# Statistical Analysis

## Primary Question: How does age group correlate with drug usage?

* Purpose: Summarize drug consumption across age groups
* Approach, calculate the standard deviation, and frequency distribution of drug usage (regular vs. less-regular) within each age group.

```{r}
summary_stats <- Data %>%
  group_by(Age) %>%
  summarise(across(all_of(drugs), list(mean = mean, sd = sd)))
summary_stats
```

### Correlation Analysis

* Quantify the relationship between age and drug usage
* Approach: treat age groups as numeric and calculate the correlation between age and drug proportions for each drug

```{r}
correlation_results = list()
age_numeric = c(1, 2, 3, 4, 5, 6)  # Numeric encoding for age groups
for (drug in drugs) {
  drug_data = drugProps_df %>% filter(Drugs == drug)
  correlation_results[[drug]] = cor.test(age_numeric, as.numeric(drug_data[2:7]))
}


for (drug in drugs) {
  print(drug)
  print(correlation_results[[drug]])
}

```

```{r}
# fisher transformation
# the proportion of variance explained
r = cor(age_numeric, alc_prop$Alcohol_Prop)
r_squared = r^2
cat("Correlation (r) =", r, "R-squared =", r_squared, "\n")


```

## Proportion Confidence Intervals (Agresti-Coull Method)

* application: For each drug, compute a 95% Agresti-Coull confidence interval for the proportion of regular users across all age groups.

```{r}
drugProps = drugProps %>%
  mutate(
    n = nTotal,
    X = usageProp * nTotal,
    p_ac = (X + 2) / (n + 4),
    se_ac = sqrt((p_ac * (1 - p_ac)) / (n + 4)),
    lower = p_ac - qnorm(0.975) * se_ac,
    upper = p_ac + qnorm(0.975) * se_ac
  )
drugProps %>%
  select(Drugs, usageProp, lower, upper)

drugProps

```


## Hypothesis Testing For Single Proportion

* Test whether the proportion of alcohol users differs from a hypothesized value 
* Define the null and alternative Hypothesis.
* compute the p-value and interpret the results
* The average proportion across all drugs in the dataset is approximately 0.242

```{r}
p0 <- 0.24  # hypothesized proportion
X <- drugProps %>% filter(Drugs == "Alcohol") %>% pull(usageProp) * nTotal
n <- nTotal
phat <- X / n
z <- (phat - p0) / sqrt((p0 * (1 - p0)) / n)
p_value <- 2 * (1 - pnorm(abs(z)))  # two-tailed test
cat("Z =", z, "P-value =", p_value, "\n")

```


## Difference in Proportions Between Age Groups

```{r}
group1 = ageGroup_Prop_DFS[[1]]  # e.g., Age group 18-24
group2 = ageGroup_Prop_DFS[[6]]  # e.g., Age group 65+

p1 = group1 %>% filter(Drugs == "Alcohol") %>% pull(usageProp)
p2 = group2 %>% filter(Drugs == "Alcohol") %>% pull(usageProp)
n1 = nTotals[1]
n2 = nTotals[6]
p_pool = (p1 * n1 + p2 * n2) / (n1 + n2)
z = (p1 - p2) / sqrt(p_pool * (1 - p_pool) * (1/n1 + 1/n2))
p_value = 2 * (1 - pnorm(abs(z)))  # two-tailed test
cat("Z =", z, "P-value =", p_value, "\n")

```


## Linear Regression Analysis

* Analyze the relationship between the number of drugs consumed and age using linear regression
* Fit a linear model using lm and evaluate slope significance.

```{r}
alc_prop = alc_prop %>%
  mutate(Age_Group = as.numeric(factor(Age_Group, levels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+"))))


lm_model = lm(Alcohol_Prop ~ as.numeric(Age_Group), data = alc_prop)
summary(lm_model)
confint(lm_model)

```
> Overall, the relationship between age group and alcohol consumption proportion. The results show a statistically significant negative association (𝑝=0.0307 p=0.0307), with alcohol consumption decreasing by approximately 3.94% for each increase in age group. The model explains about 66% of the variance in alcohol consumption (𝑅2=0.66R 2=0.66).

```{r}
# Scatter plot with regression line
ggplot(alc_prop, aes(x = as.numeric(Age_Group), y = Alcohol_Prop)) +
  geom_point(color = "blue", size = 3) +  # Scatter points
  geom_smooth(method = "lm", se = TRUE, color = "red") +  # Regression line with confidence interval
  labs(
    title = "Alcohol Consumption Proportion by Age Group",
    x = "Age Group (Numeric)",
    y = "Alcohol Consumption Proportion"
  ) +
  theme_minimal() +
  scale_x_continuous(
    breaks = 1:6,  # Match numeric Age_Group to actual groups
    labels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")
  )
```


# Hypothesis Testing For Multiple Drugs

```{r}
# Define hypothesized proportion (p0)
p0 <- 0.24

hypothesis_results <- drugProps %>%
  mutate(
    SE = sqrt((p0 * (1 - p0)) / n),  # Standard error
    Z = (usageProp - p0) / SE,      # Z-statistic
    p_value = 2 * (1 - pnorm(abs(Z))), # Two-tailed p-value
    significant = ifelse(p_value < 0.05, "Yes", "No") # Significance
  ) %>%
  select(Drugs, usageProp, Z, p_value, significant)


hypothesis_results
```

```{r}
# plot for visuals

ggplot(hypothesis_results, aes(x = reorder(Drugs, -usageProp), y = usageProp)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.7) +
  geom_hline(yintercept = p0, linetype = "dashed", color = "red") +
  labs(
    title = "Observed Proportions vs. Hypothesized Value",
    x = "Drug",
    y = "Proportion of Users"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Comparison Across Age Group Visualization

```{r}
# Prepare data for faceted plot
long_data <- drugProps_df %>%
  pivot_longer(cols = starts_with("usageProp"), names_to = "Age_Group", values_to = "Proportion") %>%
  mutate(Age_Group = factor(Age_Group, labels = age))

# Faceted line plots for each drug
ggplot(long_data, aes(x = Age_Group, y = Proportion, group = 1)) +
  geom_line(color = "blue") +
  geom_point(size = 2) +
  facet_wrap(~Drugs, scales = "free_y") +
  labs(
    title = "Drug Consumption Proportions by Age Group",
    x = "Age Group",
    y = "Proportion of Consumption"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
