---
title: "Stat 240 Project Draft"
author: "Mai Tah Lee, Jon Chang, Sophie Schrag, Michitake Kuwabara"
date: "2024-12-05"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
library(tidyverse)
theme_set(theme_minimal())
```

```{r include=FALSE}
data = read_csv("drug_consumption.csv")
```

# Stat 240 Final Project

## Introduction
Drug usage and overdose are critical public health challenges, nearly less than 100,000 people die annually from drug overdoses in the United States (“Drug Overdose Death Statistics [2023]”). The patterns of drug consumption across different age groups can help us better understand the context behind these numbers, which could further guide prevention and intervention efforts. This analysis explores how age groups correlates with the usage of various drugs, both legal (e.g., alcohol) and illegal (e.g., cannabis, ecstasy) and what the trend in different drug consumption proportions across different age groups. By exploring these relationships, we aim to identify which drugs are most commonly used within each age group and examine the characteristics associated with these trends. Ultimately, this analysis finds that the most consumed drugs were caffeine, chocolate (which can cause abnormal behaviors and other sensations that can be considered an addictive substance [“Chocolate: Food or Drug? - University of Arizona”]), and alcohol among individuals aged 18-24, and overall drug consumption decreases significantly with increased age.

## Background

In order to answer our question, we need to know how much alcohol is being consumed by each age group from age 18 to 65 and up. The age groups consist of 18-24, 24-35, 35-44, 45-54, 55-64 and 65+. Our data set consists of data from one source that had this information 

#### Age and Drug Use Data(Drug_consumption.csv)
Our data came from a Drug consumption dataset (Fehrman, Egan, and Mirkes 2015). This dataset had information on many different demographics and their consumption of various drug types, some legal and some illegal. The raw data contains many columns and almost 2000 rows (individuals); however for this study we only used columns that corresponded to age group (Age) and alcohol consumption (Alcohol). 
The Age column quantified the ages of all individuals so that each row or individual had a value across a broad range between -0.95 and 2.59,  that would put them in a certain age group. For example the value -0.95197 puts that individual into the age group 18-24 where 1.09449 puts them into 45-59. 
The Alcohol column placed each individual into one of 7 classes based on the frequency of their consumption. The classes included:

-- CLO: never used

-- CL1: used over a decade ago

-- CL2: used in the last decade

--  CL3: used in the last year

-- CL4: used in the last month

-- CL5: used in the last week

-- CL6 used in the last day. 

#### Combined Data
These data can be joined together in a separate table, grouping individuals by their respective age group, and then a statistical analysis can be conducted to see if alcohol consumption varies among age groups. 

```{r echo=F }
drugs = c("Alcohol", "Amphet", "Amyl", "Benzos", "Caff", "Cannabis", "Choc", "Coke", "Crack", "Ecstasy", "Heroin", "Ketamine", "Legalh", "LSD", "Meth","Mushrooms", "Nicotine", "Semer", "VSA")

age = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")

# Adjust and initialize variable values

Data = data %>% 
  mutate(
    
### Initialize Age
    Age = case_when(
      Age == -0.95197 ~ "18-24",
      Age == -0.07854 ~ "25-34",
      Age == 0.49788 ~ "35-44",
      Age == 1.09449 ~ "45-54", 
      Age == 1.82213 ~ "55-64", 
      Age == 2.59171 ~ "65+"
    ),
    
### Make Gender Binary (clean the data); 
# **** 0==Male(-0.48246), 1==Female (0.48246) 
    Gender = case_when(
      Gender >= 0 ~ 1, # female
      Gender < 0 ~ 0   # male
    ),
    
### Initialize Education
    Education = case_when(
      Education == -2.43591 ~ "Left School Before 16 years", 
      Education == -1.73790 ~ "Left School at 16 years", 
      Education == -1.43719 ~ "Left School at 17 years", 
      Education == -1.22751 ~ "Left School at 18 years", 
      Education == -0.61113 ~ "Some College,No Certificate Or Degree", 
      Education == -0.05921 ~ "Professional Certificate/ Diploma", 
      Education == 0.45468 ~ "University Degree", 
      Education == 1.16365 ~ "Masters Degree", 
      Education == 1.98437 ~ "Doctorate Degree", 
    ),

### Initialize Country
    Country = case_when(
      Country == -0.09765 ~"Australia",
      Country == 0.24923 ~"Canada",
      Country == -0.46841 ~"New Zealand",
      Country == -0.28519 ~"Other",
      Country == 0.21128 ~"Republic of Ireland",
      Country == 0.96082 ~"UK",
      Country == -0.57009 ~"USA"
    ),

### Initialize Ethnicity
    Ethnicity = case_when(
      Ethnicity == -0.50212 ~ "Asian",
      Ethnicity == -1.10702 ~ "Black",
      Ethnicity == 1.90725 ~ "Mixed-Black/Asian",
      Ethnicity == 0.12600 ~ "Mixed-White/Asian",
      Ethnicity == -0.22166 ~ "Mixed-White/Black",
      Ethnicity == 0.11440 ~ "Other",
      Ethnicity == -0.31685 ~ "White"
    )

# The Behavior score values will not be initialized, their numbers
# are standardized based on overall variance. (It will also be time consuming to relabel the values.)

  )

Data <- Data %>% 
  mutate(
    across(all_of(drugs), ~ case_when(     
      . == "CL0" ~ 0,
      . == "CL1" ~ 0,
      . == "CL2" ~ 0,
      . == "CL3" ~ 0,
      . == "CL4" ~ 1,
      . == "CL5" ~ 1,
      . == "CL6" ~ 1
    ))
  ) %>% select( # removing the behavioral scores
    -Nscore,
    -Escore, 
    -Oscore,
    -Ascore,
    -Cscore,
    -Impulsive,
    -SS
  )

nTotal = nrow(Data) # total amount of rows or observations of persons

usage_prop = c() # list the the proportions of each drug used within the dataset
for (val in drugs) {
  drugSum = sum(Data[[val]])
  drugProp = drugSum / nTotal
  usage_prop = append(usage_prop, drugProp)
}

# The overall proportions across all age groups
drugProps = tibble(
  Drugs = drugs,
  usageProp = usage_prop
) %>% arrange(desc(usageProp))


# The bar plot of the proportions of drug usage across all age groups
plot = ggplot(drugProps, aes(x = Drugs, y = usageProp)) + 
  geom_bar(stat="identity",fill='blue', alpha=0.5)+
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(
    title="Proportion of Drug Consumption Across All Age Groups",
    x = "Drug Consumed",
    y = "Proportion of Drug Consumption"
  )

print(plot)
```

The bar graph shows that caffeine, alcohol, and chocolate are the most consumed substances, indicating their widespread use due to accessibility and social norms. Illegal drugs like crack and heroin have much lower proportions.


```{r echo=FALSE}
#Initialize data
drugusage = data

drug_mod = drugusage %>%
  
  #Restructure dataframe for only age group and alcohol usage
  select(Age, Alcohol) %>%
  
  group_by(Age) %>% 
  
  mutate(
    
    usage_score = case_when(
      Alcohol == "CL0" ~ "less_regular",
      Alcohol == "CL1" ~ "less_regular",
      Alcohol == "CL2" ~ "less_regular",
      Alcohol == "CL3" ~ "less_regular",
      Alcohol == "CL4" ~ "regular",
      Alcohol == "CL5" ~ "regular",
      Alcohol == "CL6" ~ "regular",
    )
  ) %>% 
  
  #Summarize regular vs less regular drinkers
  group_by(Age, usage_score) %>% 
  summarize(
    n = n()
  ) %>% 
  
  pivot_wider(names_from = usage_score, values_from = n) %>%
  
  ungroup() %>%

  
  mutate(
    
    #Add total number of cases
    n = less_regular + regular,
    
    #Add usage proportion per age group
    usage_prop = regular/n,
    
    #Simplify age ranges
    Age = case_when(
      Age == -0.95197 ~ "18-24",
      Age == -0.07854 ~ "25-34",
      Age == 0.49788 ~ "35-44",
      Age == 1.09449 ~ "45-54",
      Age == 1.82213 ~ "55-64",
      Age == 2.59171 ~ "65+"
    )
  )
```

## Analysis
#### Initial Impressions

```{r echo=FALSE, warning = FALSE}
ggplot(drug_mod, aes(x = Age, y = usage_prop, group = 1)) + 
  geom_line(aes(color = "Alcohol Proportion"), linewidth = 1) + 
  geom_point(aes(color = "Alcohol Proportion"), size = 3) + 
  geom_smooth(
    method = "lm",
    aes(color = "Trend Line"), 
    linewidth = 1
  ) +
  scale_color_manual(
    values = c("Alcohol Proportion" = "blue", "Trend Line" = "red"),
    name = "Legend"
  ) +
  labs(
    title = "Alcohol Consumption Proportion by Age Group",
    x = "Age Group",
    y = "Proportion of Alcohol Consumption"
  )
```

From the graph we can see a clear and intuitive downward trend in alcohol consumption as age increases, but we will have to do a statistical analysis to say this conclusively.

## Statistical Model

We wish to explore if the average proportion of drinkers changes between age groups. To do this we will let: 

- $X_i$ represent the statistical model of regular drinkers for $i$ age group.
- $n_i$ represent the total number of cases for each age group.
- $p_i$ represent the true proportion of drinkers for each age group.
- $i$ represent the the groups with $1$ being ages 18-24, $2$ being 25-34, etc.

and we will use these models.

$$
X_1 \sim \text{Binom}(n_1, p_1) \\
X_2 \sim \text{Binom}(n_2, p_2) \\
X_3 \sim \text{Binom}(n_3, p_3) \\
X_4 \sim \text{Binom}(n_4, p_4) \\
X_5 \sim \text{Binom}(n_5, p_5) \\
X_6 \sim \text{Binom}(n_6, p_6) 
$$
We will assume that the BINS assumptions are met for each model as such:
- B: Each case is binomial, either a regular drinker or a non-regular drinker.
- I: Each case is independent from each other.
- N: The number of cases is predetermined.
- S: The probability for each case remains the same.

The assumption of independence is could be potentially not met since drinking tends to be a social activity, thus one regular drinker could affect another person.

## Hypothesis

* With these models we will begin conducting the testing of the hypothesis

$$
H_0 : X_1 - X_2 \text{ through } X_6 = 0
$$
$$
H_a : X_1 - X_2 \text{ through } X_6 > 0
$$
using a two-sample z-test for a difference in proportions.


```{r echo=FALSE}
#Testing of hypotheses

x = drug_mod$regular
n = drug_mod$n
x1 = x[1]
n1 = n[1]

# Sample proportions of drinkers 

pbar = (x1 + x[2:6]) / (n1 + n[2:6])

# Calculating test statistic for each difference in proportions

numerator = (x1 / n1) - (x[2:6] / n[2:6]) #p-hat[1] - p-hat[2 to 6]

denominator = sqrt( ((pbar * (1 - pbar)) / n1) + ((pbar * (1 - pbar)) /  n[2:6]) )

test_stat = numerator/denominator



# Calculating p-values

p_values = tibble(Age_comparison = c("18-24 to 25-34", "18-24 to 35-44", "18-24 to 45-54", "18-24 to 55-64", "18-24 to 65+"), p_value = 2*(1 - pnorm(test_stat, mean = 0, sd = 1)))

p_values = p_values %>% 
  mutate(
    significant = p_value < 0.05
  )

```


## Interpretation

> We fail to find significant evidence that the probability of people consuming alcohol is different between age group 18-24 and each other age group, EXCEPT age group 65+ (p = 0.00942), using a standard significance score of p = 0.05. However, for age group 54-65, there is some potential significance because the p-score was 0.06056 


# Discussion

The analysis above suggests that all age groups drink the same amount of alcohol, except for people who are 65 and older. 

Multiple limitations due to the data were identified in our analysis. First, the sampled population had a few biases, namely education and racial ones. Over 90% of participants reported being white, and 59.5% of participants had at least a degree or professional certification. The data also reported the age of participants in ranges, as noted in the background section, as opposed to specific ages, and limited the data to people 18 and older, potentially hiding any trends for any age group more specific than the given ranges. 

Two assumptions of the binomial model were also not met, those being the “same probability” assumption, and the “set sample size” assumption. Whether or not each individual had the same probability of consuming alcohol was the focus of the analysis, and so could not be an assumption when selecting the model. As for the sample size, the data was collected through an open online survey, the sample size was not fixed, and just depended on how many people took the survey before it was closed.

Additional analysis could be done investigating whether any of the other drugs with data reported for them have an age bias, or whether trends in any of the other participant characteristics; such as gender, education, or ethnicity; show a trend. 


# References

1. “Chocolate: Food or Drug? - University of Arizona.” N.p., n.d. Web. 4 Dec. 2024. <https://experts.arizona.edu/en/publications/chocolate-food-or-drug>.

2. “Drug Overdose Death Statistics [2023]: Opioids, Fentanyl & More.” NCDAS. N.p., n.d. Web. 2 Dec. 2024, https://drugabusestatistics.org/drug-overdose-deaths/.

3. Fehrman, Elaine, Vincent Egan, and Evgeny Mirkes. "Drug Consumption (Quantified)." UCI Machine Learning Repository, 2015, https://doi.org/10.24432/C5TC7S. 

