---
title: "STAT 240: Project Draft"
author: "Jon, Mai Tah, Michitake, Sophie"
date: "Fall 2024"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      error = FALSE, fig.height = 4)
library(tidyverse)
source("../../scripts/ggprob.R")

theme_set(theme_minimal())

```

```{r, include=FALSE}
drugusage = read.csv("../../data/drug_consumption.csv")

drug_mod = drugusage %>%
  select(Age, Alcohol) %>%
  group_by(Age, Alcohol) %>%
  summarize(
      n = n()
  ) %>% pivot_wider(
    names_from = Alcohol, values_from = n, values_fill = 0
  ) %>% 
  relocate(
    CL1, .after = CL0
  ) %>% 
  mutate(
    n = CL0 + CL1 + CL2 + CL3 + CL4 + CL5 + CL6,
    usage_score = (CL1 * 1) + (CL2 * 2) + (CL3 * 3) + (CL4 * 4) + (CL5 * 5) + (CL6 * 6)
  )
   
drug_mod2 = drugusage %>%
  select(Age, Alcohol) %>%
  group_by(Age, Alcohol) %>% 
  mutate(
    usage_score = case_when(
      Alcohol == "CL0" ~ 0,
      Alcohol == "CL1" ~ 1,
      Alcohol == "CL2" ~ 2,
      Alcohol == "CL3" ~ 3,
      Alcohol == "CL4" ~ 4,
      Alcohol == "CL5" ~ 5,
      Alcohol == "CL6" ~ 6,
    )
  )

summary(lm(formula = usage_score ~ Age, data = drug_mod))
```


# Statistical Analysis

## Initial Perspective

* First, we will set up our data by creating a variable "usage_score" to represent the usage level of a person, and we will use this variable for all future analysis.

```{r echo=FALSE}
ggplot(drug_mod2, aes(Age, usage_score)) +
  geom_point() +
  geom_smooth(color = "red", se = F, method = lm) +
  labs(title = "Usage Score by Age Group", x = "Age Group", y = "Usage Score") +
  theme_bw()

```

* Initially, the graph shows the two variables to be negatively correlated; however with this graph it is not very clear.

```{r echo=FALSE}

ggplot(drug_mod, aes(x = Age, y = usage_score)) +
  geom_point() +
  geom_smooth(color = "red", se = F, method = lm) + 
  labs(title = "Total Usage Score by Age Group", x = "Age Group", y = "Total Usage Score") +
  theme_bw()
```

* By finding the total usage score and plotting the total usage score per age group, the negative correlation becomes much more obvious.
