---
title: "STAT 240: Project Draft"
author: "Jon, Mai Tah, Michitake, Sophie"
date: "Fall 2024"
output: rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      error = FALSE, fig.height = 4)
library(tidyverse)
source("../../scripts/ggprob.R")

theme_set(theme_minimal())

```

```{r, include=FALSE}
#Initialize data
drugusage = read.csv("../../data/drug_consumption.csv")


drug_mod = drugusage %>%
  
  #Restructure dataframe for only age group and alcohol usage
  select(Age, Alcohol) %>%
  
  
  group_by(Age) %>% 
  
  mutate(
    
    usage_score = case_when(
      Alcohol == "CL0" ~ "less_regular",
      Alcohol == "CL1" ~ "less_regular",
      Alcohol == "CL2" ~ "less_regular",
      Alcohol == "CL3" ~ "less_regular",
      Alcohol == "CL4" ~ "regular",
      Alcohol == "CL5" ~ "regular",
      Alcohol == "CL6" ~ "regular",
    )
  ) %>% 
  
  #Summarize regular vs less regular drinkers
  group_by(Age, usage_score) %>% 
  summarize(
    n = n()
  ) %>% 
  
  pivot_wider(names_from = usage_score, values_from = n) %>%
  
  ungroup() %>%

  
  mutate(
    
    #Add total number of cases
    n = less_regular + regular,
    
    #Add usage proportion per age group
    usage_prop = regular/n,
    
    #Simplify age ranges
    Age = case_when(
      Age == -0.95197 ~ "18-24",
      Age == -0.07854 ~ "25-34",
      Age == 0.49788 ~ "35-44",
      Age == 1.09449 ~ "45-54",
      Age == 1.82213 ~ "55-64",
      Age == 2.59171 ~ "65+"
    )
  )


```
# Introduction

# Background

# Statistical Analysis

## Initial Thoughts

* First, we will set up a graph to take a look at the proportion of people that drink within each age group.

```{r echo=FALSE}
ggplot(drug_mod, aes(y = usage_prop, x = Age)) +
  geom_bar(stat = "identity", fill = "lightblue", color = "darkgray") + 
  labs(title = "Proportion of age group that drinks", y = "Proportion", color = "Age group") + theme_minimal()
```

* The graphs show that the proportion decreases as the age group gets older, but we will conduct further analysis to make this conclusion more accurately.

<!-- * What is our estimate of the difference between the probability a person drinks regularly while young (18-24) and while in a different age group? -->

## Statistical Model

We wish to explore if the average proportion of drinkers changes between age groups. To do this we will let: 

- $X_i$ represent the statistical model of regular drinkers for $i$ age group.
- $n_i$ represent the total number of cases for each age group.
- $p_i$ represent the true proportion of drinkers for each age group.
- $i$ represent the the groups with $1$ being ages 18-24, $2$ being 25-34, etc.

and we will use these models.

$$
X_1 \sim \text{Binom}(n_1, p_1) \\
X_2 \sim \text{Binom}(n_2, p_2) \\
X_3 \sim \text{Binom}(n_3, p_3) \\
X_4 \sim \text{Binom}(n_4, p_4) \\
X_5 \sim \text{Binom}(n_5, p_5) \\
X_6 \sim \text{Binom}(n_6, p_6) 
$$
We will assume that the BINS assumptions are met for each model as such:
- B: Each case is binomial, either a regular drinker or a non-regular drinker.
- I: Each case is independent from each other.
- N: The number of cases is predetermined.
- S: The probability for each case remains the same.

The assumption of independence is could be potentially not met since drinking tends to be a social activity, thus one regular drinker could affect another person.

## Hypothesis

* With these models we will begin conducting the testing of the hypothesis

$$
H_0 : X_1 - X_2 \text{ through } X_6 = 0
$$
$$
H_a : X_1 - X_2 \text{ through } X_6 > 0
$$
using a two-sample z-test for a difference in proportions.

```{r}
#Testing of hypotheses

drug_mod

x = drug_mod$regular
n = drug_mod$n
x1 = x[1]
n1 = n[1]

# Sample proportions of drinkers 

pbar = (x1 + x[2:6]) / (n1 + n[2:6])


# Null distribution

gnorm(0,1)


# Calculating test statistic for each difference in proportions

numerator = (x1 / n1) - (x[2:6] / n[2:6]) #p-hat[1] - p-hat[2 to 6]

denominator = sqrt( ((pbar * (1 - pbar)) / n1) + ((pbar * (1 - pbar)) /  n[2:6]) )

test_stat = numerator/denominator
test_stat

# Test statistic on null distribution
gnorm(0,1) +
  geom_vline(xintercept = test_stat, color = "red")

# Calculating p-values

p_values = tibble(Age_comparison = c("18-24 to 25-34", "18-24 to 35-44", "18-24 to 45-54", "18-24 to 55-64", "18-24 to 65+"), p_value = 2*(1 - pnorm(test_stat, mean = 0, sd = 1)))

```


## Interpretation

> We fail to find significant evidence that the probability of people consuming alcohol is different between age group 18-24 and each other age group, EXCEPT age group 65+ (p = 0.00942), using a standard significance score of p = 0.05. However, for age group 54-65, there is some potential significance because the p-score was 0.06056 

# Discussion